---
title: "Expression array normalization and analysis"
author: "Lindsay A Holden"
date: "June 6, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Outline

Initial normalization based on Mattick Lab protocol, see  
<http://matticklab.com/index.php?title=Single_channel_analysis_of_Agilent_microarray_data_with_Limma>  

1. normalize arrays  

    a) load libraries, target file, annotations, and gene list load data into an RGList object  
    b) assign spot types and set the status of each spot on the array  
    c) subract background fluorescence  
    d) normalize between arrays  
    e) average replicate spots  
 \ 
2. applying linear model  

    a) create a design matrix for linear modeling function  
    b) apply intensity values to lmFit (linear model)  
    c) create contrast matrix for your comparison of interest  
    d) apply contrast matrix to modeled data & compute statistics  
    e) output stats  
    f) annotate output  

## Normalize arrays

### Load libraries

Load the libraries you'll be using and set your working directory (where you want R to look for files and to save files).

```{r}
library(limma)
setwd("E:/LINDSAY/expression arrays/all arrays")
```

### Load the arrays

Do this by first

