---
title: "Expression_array_normalization_and_analysis"
author: "Lindsay A Holden"
date: "June 7, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "E:/LINDSAY/expression arrays/all arrays") ## working directory for all R chunks
```

## Outline

Initial normalization based on Mattick Lab protocol, see  
<http://matticklab.com/index.php?title=Single_channel_analysis_of_Agilent_microarray_data_with_Limma>  

1. normalize arrays  

    a) load libraries, target file, annotations, and gene list  
    b) load data into an RGList object  
    c) assign spot types and set the status of each spot on the array  
    d) subract background fluorescence  
    e) normalize between arrays  
    f) average replicate spots  
 \ 
2. applying linear model  

    a) create a design matrix for linear modeling function  
    b) apply intensity values to lmFit (linear model)  
    c) create contrast matrix for your comparison of interest  
    d) apply contrast matrix to modeled data & compute statistics  
    e) output stats  
    f) annotate output  

## Reading Microarray Data

### Load libraries

Load the libraries you'll be using.

```{r libraries}
library(limma)
```

### Load target file

Load your target file containing SampleNumber, FileName, and Condition data. Make sure to include full filepath. In my target file each row is one sample and each column is one descriptor.

```{r targets}
targets <- readTargets(file = "targets.txt")
names(targets) ## to view the column names
```

### Load annotations

These annotations are what come from Agilent. Currently, they are Zv9 (danRer7) and should be queried against more updated releases of RefGene, ENTREZ, ENSEMBL, etc.

```{r annotations}
annotations <- read.csv("AllAnnotations.csv")
names(annotations) ## to view the column names
```

### Load genelist

The genelist is pretty much the same thing as annotations and also come from Agilent. Currently, they are Zv9 (danRer7) and should be queried against more updated releases of RefGene, ENTREZ, ENSEMBL, etc. 

```{r genelist}
genelist <- read.delim2("E:/LINDSAY/expression arrays/all arrays/genelist_withcontrols.txt")
names(genelist) ## to view the column names 
```

### Image-derived spot quality weights

First, define which spots you want to weight. Do this by writing a function that weights strongly positive spots as 1 and flagged spots as 0. We will apply this when we load the array files.

```{r}
myFlagFun <- function(x) {
#Weight only strongly positive spots 1, everything else 0
present <- x$gIsPosAndSignif == 1
probe <- x$ControlType == 0
manual <- x$IsManualFlag == 0
strong <- x$gIsWellAboveBG == 1
y <- as.numeric(present & probe & manual & strong)
#Weight flagged spots 0
sat <- x$gIsSaturated == 0
xdr <- x$gIsLowPMTScaledUp == 0
featureOL1 <- x$gIsFeatNonUnifOL == 0
featureOL2 <- x$gIsFeatPopnOL == 0
flagged <- (sat & xdr & featureOL1 & featureOL2)
flagged <- grep(FALSE, flagged)
good <- grep(TRUE, y==1)
flagged <- intersect(flagged, good)
y[flagged] <- 0
y
}

```


### Load array files

Using read.maimages, load extracted array files (txt output from Agilent's Feature Extraction Program) as an EListRaw from a set of one-color files.

```{r data, results = "hide"}
a <- read.maimages(targets,
                   source = "agilent", ## same as "agilent.median" other option is "agilent.mean"
                   green.only = TRUE,
                   wt.fun = myFlagFun)
```


This takes a minute or two (depending on how many files you have to load). Then check how many arrays loaded.

```{r data dimensions}
dim(a)
colnames(a)
```

### Reading probe annotation

a$genes had a ProbeID vector, but it only contains numbers at this point. We want to add human-readable probe IDs. These are found in the .gal file provided in the Agilent design file (026437). I dragged and dropped the .gal file into my directory.

```{r}
a$genes <- readGAL(galfile = "026437_D_20140627.gal",
                   path = "C:/Users/holdenl/Documents/ExpressionArrays")
dim(a$genes)
names(a$genes)
```

### Printer layout

This is already loaded into our EListRaw. We have commerically produce Agilent arrays, so this shouldn't really matter, but we'll look at it anyway.

```{r}
names(a$printer)
```


### Assign spot types

The SpotType file allows you to identify spots when visualizing sampels on plots to assess sample quality. First load the file, see what it contains, then assign the status of each spot on the array.

```{r spottypes}
spottypes <- readSpotTypes(file = "SpotType.txt",
                           sep = "\t")
names(spottypes)
a$genes$Status <- controlStatus(spottypes, a)
```

## Quality Assessment

### Visualize on an MA-plot

For single channel data objects, a between-array MD-plot is produced. An articifial array is produced by averaging all the arrays other than the array specified. A mean-difference plot is then producing from the specified array and the artificial array. Note that this procedure reduces to an ordinary mean-difference plot when there are just two arrays total. 

```{r}
plotMD(a, legend = FALSE) ## uses the first array file vs all others
plotMA3by2(a, path = "C:/Users/holdenl/Documents/ExpressionArrays/raw_MAplots") ## all arrays individually vs all others...will take a wee bit and prints out png files
```

### Visualize background intensity boxplots

This allows for you to see if you have any weird outliers with abnormal background. If we color by "run", then we can see that 1st (black), 2nd (red), or 3rd (green) run aren't unequally affected by background spread.

```{r}
boxplot(data.frame(log2(a$Eb)), col = c(1,2,3), group = targets$run, outline = FALSE, main = "Bkgd by run", ylab = "log2 bkgd")
```

If we color by "strain", then we can see that AB (black), TU (red), or WIK (green) run aren't unequally affected by background spread.

```{r}
boxplot(data.frame(log2(a$Eb)), col = c(1,2,3), group = targets$treatment, outline = FALSE, main = "Bkgd by treatment", ylab = "log2 bkgd")
```

### Spatial heterogeneity

```{r}
#imageplot(log2(a$Eb[,1]),a$printer)
```


